Dirace: Practical Proof Automation of Dirac Notation Equations

### 2025-01-20 flowchart.tex
2025/1/17
2025/1/20

    % \node (rename) [process] {rename unique bound variable names};
    \node (rewrite1) [rewriting] {first rewritings};
    \node (expansion) [process, below of=rewrite1] {variable expansion};
    \node (rewrite2) [rewriting, below of=expansion] {second rewritings};
    \node (sorting) [process, below of=rewrite2] {sorting modulo bound variables};
    \node (successive) [process, below of=sorting] {sorting successive summation};
    \node (deBruijn) [process, below of=successive] {de Bruijn normalization};

    % \draw [arrow] (rename) -- (rewrite1);
    \draw [arrow] (rewrite1) -- (expansion);
    \draw [arrow] (expansion) -- (rewrite2);
    \draw [arrow] (rewrite2) -- (sorting);
    \draw [arrow] (sorting) -- (successive);
    \draw [arrow] (successive) -- (deBruijn);

>>>>

    % \node (rename) [process] {rename unique bound variable names};
    \node (rewrite1) [rewriting] {first rewritings};
    \node (expansion) [process, below of=rewrite1] {variable expansion};
    \node (rewrite2) [rewriting, below of=expansion] {second rewritings};
    \node (sort) [process, below of=rewrite2] {sorting without bound variables};
    \node (swap) [process, below of=sort] {swap successive summation};
    \node (deBruijn) [process, below of=swap] {de Bruijn normalization};

    % \draw [arrow] (rename) -- (rewrite1);
    \draw [arrow] (rewrite1) -- (expansion);
    \draw [arrow] (expansion) -- (rewrite2);
    \draw [arrow] (rewrite2) -- (sort);
    \draw [arrow] (sort) -- (swap);
    \draw [arrow] (swap) -- (deBruijn);

### 2025-01-20 labelled.tex
2025/1/20

Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators. This allows us to talk aboud the states and operations locally, without referring to the whole system. For instance, assume $Q$ and $R$ are two registers, we have
\[
    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \ket{\Phi})_{(Q, R)}.
\]
In the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.

@@

We assume the set of register symbols is $\cR$. The syntax of quantum registers is defined as follows.
\begin{definition}[quantum registers]
  \begin{align*}
    R ::= r\in\cR \mid (R, R)
  \end{align*}
\end{definition}

Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notations.

To reason about the registers, we define their variable set as the enumeration of all register symbols involved.

\begin{definition}[register variable set]
The variable set of a register is defined inductively by:
\begin{itemize}
    \item $R\equiv r$: $\var(R) = \{r\}$;
    \item $R\equiv (R_1,R_2)$: $\var(R) = \var(R_1) \cup\var(R_2)$.
\end{itemize}
\end{definition}

\textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
% \begin{itemize}
$ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
% \end{itemize}

@@

\subsubsection*{Typing rules}

@@

The typing rule for tensor product is
\[
    \frac{
        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
        \bigcap_i s_i = \emptyset \qquad
        \bigcap_i s_i' = \emptyset
    }
    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\]

The typing rule for dot product is
\[
    \qquad
    \frac{
        \begin{aligned}
            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
        \end{aligned}
        \qquad 
        \begin{aligned}
            s_1 \cap s_2 \backslash s_1' = \emptyset \\
            s_2' \cap s_1' \backslash s_2 = \emptyset
        \end{aligned}
    }
    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
\]

@@

\begin{align*}
    & \textrm{(R-SUM-PUSHD0)}
    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    %
    & \textrm{(R-SUM-PUSHD1)}
    && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
\end{align*}

\begin{align*}
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m) \\
    %
\end{align*}

\subsubsection*{Eliminating the registers}

\begin{enumerate}
  \item For any register $R$ s.t. $E[\Gamma]\vdash R : \reg \sigma$, suppose $\var R = \{r_1,r_2,\cdots,r_n\}$, 
  we introduce variables $i_{r_k} : \Basis(\sigma_{r_k})$ with $r_k : \reg \sigma_{r_k} \in E$ for $k = 1,\cdots, n$. We reconstruct the basis $|i_R\>$ (which is of type $\KType(\sigma)$) and $\<i_R|$ (which is of type $\BType(\sigma)$) of $R$ by:
  \begin{itemize}
    \item $R = r_k$, $|i_R\> \triangleq |i_{r_k}\>$ and $\<i_R| \triangleq \<i_{r_k}|$;
    \item $R = (R_1,R_2)$: $|i_R\> \triangleq |i_{R_1}\> \otimes |i_{R_2}\>$ and $\<i_R| \triangleq \<i_{R_1}| \otimes \<i_{R_2}|$.
  \end{itemize}
    \item Expansion of all $K_R,B_R,O_{R,R'}$

    After the rewriting of this step, the expression we have is always in the following form:
    $$
    \Big(\sum_{i_1}\cdots\sum_{i_n} a.\tD_f\Big) + \cdots + \Big(\sum_{i_1}\cdots\sum_{i_{n'}} a.\tD_f\Big)
    $$
    where $\tD_f$ is generated by :
    $$\tD_f ::= |i\>_r \mid {}_r\<i| \mid \te\otimes\te \mid \te\cdot\te. $$
    \item
    Suppose $\mathsf{CDom}(\tD_{f}) = \{(i_1,r_1),\cdots,(i_n,r_n)\}$ and $\mathsf{Dom}(\tD_{f}) = \{(i'_1,r'_1),\cdots,(i'_{n'},r'_{n'})\}$ where the list is sorted by a default order of variables in $\cR$. We define:
    $$D_f = \bigg(\prod_{(i,j)\in \mathsf{Can}(\tD_f)}\delta_{i,j}\bigg).\Big((|i_1\>\otimes\cdots\otimes|i_n\>) \cdot (\<i'_1|\otimes\cdots\otimes\<i'_{n'}|)\Big).$$
    If $\mathsf{Dom}$ or $\mathsf{CDom}$ is an empty list, the $D_f$ is just a Ket or Bra.
    Then we claim:
    \begin{itemize}
        \item If $E[\Gamma]\vdash \tD_f : \DType(s_1,s_2)$, then $\{r_1,\cdots,r_n\} = s_1$ and
        $\{r'_1,\cdots,r'_{n'}\} = s_2$.
        \item Set $R = ((\cdots(r_1,r_2),\cdots),r_n)$ and $R' = ((\cdots(r'_1,r'_2),\cdots),r'_{n'})$, then
        $$\tD_{f} = (D_f)_{R,R'}$$
    \end{itemize}
    \item For the obtained formula in (3), we replace each $\tD_f$ by $D_f$ which is computed by step (4). Note that $D_f$ is a Dirac term without labels, so the equivalence of two labelled Dirac terms is reduced to decide the equivalence of two Dirac terms. This is guaranteed if two labelled Dirac terms are of the same type (by typing rules).
\end{enumerate}

>>>>

Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators. This allows us to talk aboud the states and operations locally, without referring to the whole system. For instance, assume $Q$ and $R$ are two registers, we have
\[
    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \ket{\Phi})_{(Q, R)}.
\]
On the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$. It is equivalent to extend the operation using identity operators on other subsystems (i.e. the cylinder extension), and consider the application in the whole system.

@@

We assume the set of register symbols is $\cR$. The syntax of quantum registers is defined as follows.
\begin{definition}[quantum registers]
  \begin{align*}
    R ::= r\in\cR \mid (R, R).
  \end{align*}
\end{definition}

Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notations.
To reason about the registers, we define their variable set as the enumeration of all register symbols involved.

\begin{definition}[register variable set]
The variable set of a register is defined inductively by:
\begin{itemize}
    \item $\var(R) = \{r\}$, if $R\equiv r$; or
    \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
\end{itemize}
\end{definition}

% \textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
% % \begin{itemize}
% $ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
% \end{itemize}

@@

\subsubsection*{Typing rules}
Some typing rules are introduced here.
The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.

@@

The dot and the tensor product symbols are different from those in unlabelled Dirac notation. Since the goal of labels is to replace the order and structure of tensor products by the reference to registers, the tensor product becomes an AC symbol. The typing still checks whether the component subsystems are disjoint with each other.
\[
    \frac{
        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
        \bigcap_i s_i = \emptyset \qquad
        \bigcap_i s_i' = \emptyset
    }
    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
\]
As for the dot product, the disjointness is considered except registers contracted by multiplication.
\[
    \qquad
    \frac{
        \begin{aligned}
            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
        \end{aligned}
        \qquad 
        \begin{aligned}
            s_1 \cap s_2 \backslash s_1' = \emptyset \\
            s_2' \cap s_1' \backslash s_2 = \emptyset
        \end{aligned}
    }
    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
\]

@@

This step reduces all terms in the form of $K_R$ or $O_{R;R}$.
Other symbols on labelled Dirac notations are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.

The final part of rules operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.

\begin{align*}
    & \textrm{(R-SUM-PUSHD0)}
    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
    %
    & \textrm{(R-SUM-PUSHD1)}
    && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
    %
    & \textrm{(R-L-SORT0)}
    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
    %
    & \textrm{(R-L-SORT1)}
    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
    %
    & \textrm{(R-L-SORT2)}
    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m) \\
    %
\end{align*}

These rules are added to the rewriting system in ~\Cref{sec: decide} and executed together.
In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is Dirac notation scalar with labelled basis.

\begin{lemma}
    For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
    \[
    \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
    + \cdots +
    \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
    \]
    Where $a_i$ are scalar typed Dirac notations.
\end{lemma}

### 2025-01-20 main.tex
2025/1/20

In practice, Dirac notations are combined with other syntax for better expressiveness. For example, the big operator sum $\sum_{i \in S} A$ are used in theoretical reasonings. We also use labelled Dirac notation, where Dirac notations are subscripted with registers to denote the quantum system in consideration, e.g. $\ket{i}_q \otimes \ket{i}_r$.

We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
A universal algebra defines a signature of function symbols and the axioms. Terms in the algebra are constructed by constants, variables or function application on terms. In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$, and the axioms Dirac notation equivalenc rules with equational logic.

@@

\yx{I guess it's wise to avoid mentioning that we don't impose fixed arity.}

@@

\yx{where should we put the example?}
To conclude the design introduction, we demonstrate part of the formalization of the preliminary example.
\begin{example}[formalizing the preliminary example]
    Definitions and assumptions in the environment:
    \begin{align*}
        & \textrm{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
        & && : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
        & \textrm{phi} && := \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
        & \textrm{T} && : \Index; \\
        & \textrm{M} && : \OType(T, T).
    \end{align*}
    The symbol TPO is the encoding of operator transpose. It takes two index $T_1$ and $T_2$, and transforms the operators.
    phi takes the index $T$ and defines the maximally entangled states. It sums over all basis in $T$, indicated by the universal set $\mathbf{U}(T)$.
    With the assumption of index T and operator M, we can write the equivalence as 
    \[
        (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
    \]
\end{example}

@@

\section{Algorithm for Deciding Dirac notations equations}

@@

\subsection{Normalization modulo $E$ by Term Rewriting}
Term rewriting rules, written as $l -> r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$. The procedure terminates when no more rewritings can be made, and the order of rewritings will be irrelevant if the term rewriting system is \textit{confluent}, which is a desirable property.

@@

    The relation $e_1 <_\mathcal{B} e_2$ between the two terms holds when:
    \begin{itemize}
        \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
        \item $head(e_1) < head(e_2)$, or
        \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
    \end{itemize}
\end{definition}
It can be checked that $e_1 =_\mathcal{B} e_2$ if and only if neither $e_1 <_\mathcal{B} e_2$ or $e_2 <_\mathcal{B} e_1$ holds.
The idea is to compare the function symbols in the top down order, and omit bound variable occurance.

@@

The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.

@@

Here is the code for the preliminary example.

>>>>

In practice, Dirac notations are combined with other syntax, such as the big operator sum $\sum_{i \in S} A$, for better expressiveness. We also use labelled Dirac notations like $\ket{i}_q \otimes \ket{i}_r$ to denote the quantum system in consideration, where Dirac notations are subscripted with registers.

We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
A universal algebra defines a signature of function. Terms in the algebra are constructed by constants, variables or function application on terms. In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
Reasonings of Dirac notation terms follow the equational logic, where an equaitonal relation $e_1 \doteq e_2$ is defined, consistent with substitution and term construction:
\[
    \frac{E \vdash s \doteq t}{E \vdash v(s) \doteq v(t)}
    \qquad
    \frac{E \vdash s_1 \doteq t_1 \qquad \cdots \qquad E \vdash s_n \doteq t_n}{E \vdash f(s_1, \dots, s_n) \doteq f(t_1, \dots, t_n)}
\]
Here $v(s)$ represent a substitution of variables, and $E$ is set of Dirac notation equational axioms.

@@

@@

\yx{where should we put the example?}
To conclude the design introduction, we demonstrate part of the formalization of the motivating example.
\begin{example}[formalizing the motivating example]
    Definitions and assumptions in the environment:
    \begin{align*}
        & \textrm{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
        & && : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
        & \textrm{phi} && := \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
        & \textrm{T} && : \Index; \\
        & \textrm{M} && : \OType(T, T).
    \end{align*}
    The symbol TPO is the encoding of operator transpose. It takes two index $T_1$ and $T_2$, and transforms the operators.
    phi takes the index $T$ and defines the maximally entangled states. It sums over all basis in $T$, indicated by the universal set $\mathbf{U}(T)$.
    With the assumption of index T and operator M, we can write the equivalence as 
    \[
        (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
    \]
\end{example}

@@

\section{Algorithm for Deciding Dirac notations equations}
\label{sec: decide}

@@

\subsection{Normalization modulo $E$ by Term Rewriting}
Term rewriting rules, written as $l \to r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$. The procedure terminates when no more rewritings can be made, and the order of rewritings will be irrelevant if the term rewriting system is \textit{confluent}, which is a desirable property.

@@

    The relation $e_1 <_\mathcal{B} e_2$ between the two terms holds when:
    \begin{itemize}
        \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
        \item $head(e_1) < head(e_2)$, or
        \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
    \end{itemize}
\end{definition}
It can be checked that $e_1 =_\mathcal{B} e_2$ if and only if neither $e_1 <_\mathcal{B} e_2$ or $e_2 <_\mathcal{B} e_1$ holds.
The idea is to compare the function symbols in the top down order, and omit bound variable occurance. 
With the order, the sort function normalize the term with respect to AC equivalence.

\begin{definition}[sort transformation]
    For a term $e$ with bound variable set $\mathcal{B}$,
    The sort transformation sort(e) is defined as
    \begin{itemize}
        \item $e$, if $e$ is a variable or constant;
        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$. If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
    \end{itemize}
\end{definition}

The order of bound variables can be decided according to the result of the sort transformation, and then successive summations can be normalized by swapping.
\begin{definition}[swap transformation]
    For a term $e$ with sorting result sort$(e)$, we can obtain a total order of all bound variables appearing in the term $e$. Specifically, they are ordered by their first appearances, except in function definitions $\lambda x$ and $\mu x$.
    The swap transformation then arrange successive summation according to this order.
\end{definition}

@@

% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.

@@

Here is the code for the motivating example.
